---
title: "From model output to figure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From model output to figure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6, fig.height = 5, dpi = 120, fig.align = "center"
)
library(fancymaps)

# Pinned so this vignette builds the same figures everywhere. The example grid
# is 194 km across, just under the 200 km where coastline() starts asking for a
# 1:10m shoreline -- so left alone it would draw a different coastline
# depending on which optional packages the building machine happens to have.
land <- coastline_fixture()
grid <- example_grid()
```

A spatial model ends in a handful of figures: the predicted surface, that
surface beside its uncertainty, the same surface across seasons, and the
effort that produced the data. This vignette walks that pipeline once, from
the forms model output actually arrives in to the figures a manuscript
carries, and ends with the two mistakes that the first pipeline to adopt this
package made -- both of them upstream of the package, and both worth knowing
about before they cost you an afternoon.

Everything here draws from `example_grid()`, a small `sf` grid in the Gulf of
Maine with one column of each kind of quantity these models emit, and from
`coastline_fixture()`, a bundled shoreline. Substitute your own grid and the
calls do not change.

## The forms output arrives in

Model output reaches the figure stage in one of three shapes, and
[`as_map_data()`] -- called for you by every map function -- accepts each of
them.

**An `sf` object with the predictions as columns.** The tidiest case, and the
one `example_grid()` is in:

```{r sf-form}
grid
```

**A plain data frame with coordinate columns.** What remains once geometry has
been dropped for modelling, which is most modelling. Two numeric columns say
nothing about whether they are degrees or metres, so `crs` is required rather
than guessed:

```{r df-form, message = FALSE}
proj <- equal_area_crs(grid)
centres <- sf::st_centroid(sf::st_geometry(sf::st_transform(grid, proj)))

flat <- data.frame(sf::st_coordinates(centres), density = grid$density)
head(flat, 3)

map_surface(flat, "density", coords = c("X", "Y"), crs = proj,
            label = "animals per km2", coastline = land)
```

**A named vector keyed by cell id.** The form a posterior summary or a lookup
table arrives in -- values computed elsewhere, matched back to the grid by
`by`:

```{r keyed-form, message = FALSE}
posterior_mean <- setNames(grid$density * 1.1, grid$grid_id)

map_surface(grid, posterior_mean, by = "grid_id",
            label = "posterior mean density", coastline = land)
```

The join is checked: every name must be found in the `by` column, and a
missing key stops the call rather than leaving a hole. A partial join produces
a map with gaps that looks exactly like a region where the model had nothing
to say, and nobody catches that by looking.

To see what a map will draw before it draws it, call `as_map_data()` yourself:

```{r inspect}
as_map_data(grid, "density")
```

## The deliverable

```{r surface, message = TRUE}
map_surface(grid, "density", label = "animals per km2",
            coastline = land,
            title = "Predicted density",
            caption = "Example fixture; summer period, no availability correction.")
```

Note what happened without being asked: the projection was chosen and applied
to every layer, the scale went logarithmic because the data are skewed -- and
said so, both in the console and in the legend -- and the top percentile was
capped, with the top break marked `≥` rather than the bright cells being
dropped. `vignette("scales")` is the full argument for each of those choices;
the short version is that anything the figure had to decide is reported on the
figure, where a reader can see it.

The caption is where provenance belongs -- which period, which product, which
correction was not applied -- and anything the function had to decide is
appended to it rather than replacing it.

## The value beside its uncertainty

A predicted surface without its uncertainty is half a claim. [`map_pair()`]
draws the two as one figure: same extent, same projection, same coastline,
aligned panels, furniture once.

```{r pair, message = FALSE, fig.width = 8, fig.height = 4.5}
map_pair(grid, "density", "cv",
         labels = c("animals per km2", "CV"),
         titles = c("Predicted density", "How uncertain"),
         coastline = land)
```

The legends stay separate on purpose -- density and CV are different
quantities in different units, and a shared ramp would be a lie. What the
panels share is extent, projection, position, size and typography, which is
what makes the pairing read as one figure.

When the uncertainty is a signed score rather than a spread -- an
extrapolation surface, say -- the right panel diverges instead:

```{r pair-mess, message = FALSE, fig.width = 8, fig.height = 4.5}
map_pair(grid, "density", "mess",
         uncertainty_kind = "diverging", uncertainty_direction = -1,
         labels = c("animals per km2", "MESS"),
         titles = c("Predicted density", "How familiar these conditions are"),
         coastline = land)
```

## The series

Panels drawn separately cannot be compared and look as though they can: each
one stretches its own range across the full ramp, and the bright patch in a
sparse year is drawn in the same yellow as the bright patch in an abundant
one. [`map_panels()`] puts every panel on one scale with one legend, which is
the only arrangement in which "brighter" means "more".

```{r panels, message = FALSE, fig.width = 7, fig.height = 3.2}
seasons <- cbind(spring = grid$density,
                 summer = grid$density * 2.5,
                 autumn = grid$density * 0.4)

map_panels(grid, seasons, label = "animals per km2", coastline = land)
```

The matrix form -- one row per cell, one column per period, column names
becoming titles -- is the shape an averaging or posterior step emits, so
usually nothing needs reshaping.

## The effort

The map that answers "where did you look", which reviewers ask for and
pipelines draw last:

```{r effort, message = FALSE}
set.seed(11)
midpoints <- sf::st_as_sf(
  data.frame(lon = stats::runif(200, -70.3, -68.2),
             lat = stats::runif(200, 42.7, 44.2)),
  coords = c("lon", "lat"), crs = 4326)

map_effort(points = midpoints, bin = TRUE, bins = 15, coastline = land)
```

## The two coordinate-system questions

A map involves two different CRS questions, answered by two different
functions on purpose.

**What is the map drawn in?** [`display_crs()`] decides once per figure and
applies the answer to every layer. Projected data keeps its own projection --
if the analysis was in UTM 19N, reprojecting for the figure alone would draw
something slightly other than what was fitted. Geographic data is projected to
a Lambert azimuthal equal-area centred on the data, because drawing raw
lon/lat is what makes a northern study area look stretched.

**What are areas measured in?** [`equal_area_crs()`] and [`area_km2()`], for
the abundance-by-region arithmetic, which should never be done in whatever CRS
happened to be convenient for display.

The one thing worth doing deliberately: when a manuscript carries several
maps, pass the same `crs` to all of them, for the same reason you would fix
`transform` and `limits` -- figures that are going to be compared should not
each resolve their own defaults.

## Two mistakes to make only once

Both were made by the first pipeline that adopted this package, both are
upstream of anything a map function can check, and both produced figures that
looked fine.

**Annotating a finished map with `geom_sf()` silently discards the extent.**
A returned map carries a `coord_sf()` holding the resolved projection and the
fixed extent; adding another `geom_sf` layer replaces that coordinate system
and the map quietly re-fits its limits to the union of the data. Annotate with
`geom_point()` or `geom_text()` in the display CRS's own units instead -- or,
for a study-area outline, pass `region =`, which exists so the common case
never meets this problem.

**A model frame in kilometres has to be multiplied back before mapping.**
Pipelines rescale coordinates to kilometres for numerical stability, and a
grid whose "metres" are actually kilometres draws a map one thousandth the
size -- with a scale bar labelled in units that make it look plausible,
because 400 is a believable number in either unit. Nothing downstream can
catch this; the fix is a comment at the rescaling site and a glance at the
scale bar on the first figure.

## Where to next

* `vignette("scales")` -- why the colour scales do what they do, and how to
  override each decision.
* [`leaflet_surface()`] and friends -- the same maps as interactive widgets,
  built from the same scale objects so a cell is the same colour in both
  renderers.
