---
title: "Choosing a scale"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing a scale}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6, fig.height = 5, dpi = 120, fig.align = "center"
)
library(fancymaps)

# Pinned so this vignette builds the same figures everywhere. The example grid
# is 194 km across, just under the 200 km where coastline() starts asking for a
# 1:10m shoreline -- so left alone it would draw a different coastline
# depending on which optional packages the building machine happens to have.
land <- coastline_fixture()
grid <- example_grid()
```

A map scale is where most of the information in a spatial model either arrives
or is thrown away, and it is almost always chosen by default. This vignette is
about the defaults this package uses instead, and about why each one is what it
is -- because a default you cannot argue with is a default you cannot correct.

## The problem, on a real quantity

Predicted density from a distance-sampling model is severely skewed. A handful
of cells carry most of the animals and the rest carry almost none, which is not
a defect of the model but a fact about how animals are distributed.

```{r skew}
summary(grid$density)

# how far the top of the range is above the middle
round(quantile(grid$density, 0.99) / median(grid$density))
```

The 99th percentile is over a hundred times the median. Drawn linearly, the
scale spends nearly all of its colour range on values that almost no cell has:

```{r linear}
map_surface(grid, "density", transform = "identity",
            label = "animals per km2", coastline = land,
            title = "A linear ramp on a skewed surface")
```

One bright spot on a dark field. The map is not wrong -- it is an accurate
picture of a quantity most of whose variation is at the bottom of the range,
drawn on a scale with no resolution there.

## What is wrong with `trans = "sqrt"`

The usual response is a square-root transform, and it does look better:

```{r sqrt}
map_surface(grid, "density", transform = "sqrt", label = "animals per km2",
            coastline = land, title = "The usual fix")
```

Three things are wrong with stopping here, and none of them is that the figure
looks bad.

**Nothing chose the square root.** It is the transform that happened to help.
Nothing about density says the interesting comparisons are between square roots
of it, and if the skew had been milder or worse the same call would have been
made anyway.

**The breaks are wherever the algorithm put them.** They are round numbers in
square-root space, which is not a space the reader is in.

**The legend does not say the axis is not linear.** A reader who does not
already know reads evenly spaced labels as an evenly spaced quantity, and every
comparison they make from the figure is wrong by the transform.

## What replaces it

```{r auto}
map_surface(grid, "density", label = "animals per km2", coastline = land,
            title = "Predicted density")
```

Four things changed, and all four are visible in the legend.

### The transform is chosen by a stated rule

The rule is the ratio of the 99th percentile to the median: below 10, the
values are drawn linearly; at or above 10, logarithmically. [`surface_scale()`]
does the choosing and can be called on its own to see what a map is going to
do before it does it:

```{r spec}
spec <- surface_scale(grid$density)
spec$limits
spec$breaks
```

It says so when it fires, because a default that varies with the data is a
default that has to be visible:

```{r message, echo = TRUE, message = TRUE}
invisible(surface_scale(grid$density))
```

Ten is a judgement, not a theorem. It is roughly where a linear ramp stops
having any resolution in the middle of the data, and the point of writing it
down is that you can disagree with it in one place rather than per figure.

### Zeros do not disappear

A log scale cannot show a zero and a density surface is full of them. The
transform is `log(x + offset)`, where `offset` is the 5th percentile of the
positive values:

```{r offset}
spec$note
```

`log(x + 1)` would have been simpler and is the common choice. It is also
wrong in a way that is easy to miss: adding 1 to a density in animals per km2
is adding an enormous amount, and adding 1 to the same quantity in animals per
m2 is adding almost nothing. The same code would produce two different figures
from the same data in different units. A percentile of the data has no units
problem, and the legend reports it.

### Breaks are round in the reader's units

`0.0005`, `0.002`, `0.01` -- powers of ten and their doubles and halves, in
animals per km2, which is the quantity being discussed. Not round numbers in
log space, which are round in a space nobody is looking at.

Four labels is the ceiling. It is set by the narrowest legend this package
draws, which is a horizontal colourbar under one panel of a paired figure, and
a log scale's labels are long.

### The scale is capped, and says so

Limits come from a quantile rather than the range, so one extreme cell cannot
flatten everything else. Values past the cap are **squished** to it rather than
dropped:

```{r squish}
spec$squished
max(grid$density) > spec$limits[2]
```

Dropping them would colour those cells `NA` grey -- and grey already means "the
model said nothing here", so a censored cell would be reporting the wrong thing
in a colour that already has a meaning.

Two things then say the cap is there. The top break is the cap itself, marked:

```{r labels}
spec$labels(spec$breaks)
```

and the figure carries a caption saying so, which is why the map above has one
without being given any text.

## A bounded quantity is a different problem

Occupancy probability is not a skewed positive quantity that happens to stop at
1, and [`map_probability()`] treats it as its own kind:

```{r probability}
map_probability(grid, "occupancy", label = "occupancy", coastline = land,
                title = "Posterior occupancy")
```

**The ends are fixed at 0 and 1**, not taken from the data. A map whose values
ran from 0.2 to 0.6, stretched across a full ramp, would use exactly the
colours a map running 0 to 1 uses -- and those are not the same claim. Fixed
ends also mean 0.6 is the same colour in every figure, which is what makes a
series of seasons comparable at a glance.

**The ramp is different too.** A sequential ramp puts its most saturated colour
at the top of the observed data; here the top is certainty. A cell at 0.02
should look nearly like a cell at 0, with weight arriving only as the value
approaches 1.

## A diverging quantity needs a centre, and it has no default

[`map_diverging()`] requires `midpoint`:

```{r midpoint-error, error = TRUE}
map_diverging(grid, "mess")
```

That is deliberate, and it is the one place this package refuses to guess.
[`ggplot2::scale_fill_gradient2()`] falls back to the middle of the range, which
is almost never the meaning.

For an extrapolation score, zero is the meaning: it is where inside the training
range becomes outside it.

```{r mess}
map_diverging(grid, "mess", midpoint = 0, direction = -1, label = "MESS",
              coastline = land,
              title = "How familiar these conditions are",
              subtitle = "negative: outside the range the model was fitted under")
```

`direction = -1` is there because which end of a ramp carries the visual weight
is a claim about which end matters, and it cannot be derived from the numbers.
On an extrapolation surface the values a reader has to see are the negative
ones -- they are where the model is guessing -- and they are a small minority.

### The case that makes this worth enforcing

Deviance residuals do not average zero. Centring them on zero puts almost every
cell on one side of the ramp and colours them all much the same:

```{r residual-wrong}
mean(grid$residual)

map_diverging(grid, "residual", midpoint = 0, label = "deviance residual",
              coastline = land, title = "Centred on zero, which is not the centre")
```

Centred on their own mean, the same numbers show where the model runs high and
where it runs low:

```{r residual-right}
map_diverging(grid, "residual", midpoint = mean(grid$residual),
              label = "deviance residual", coastline = land,
              title = "Centred on the mean of the residuals")
```

Both figures are of the same data. The first is the one you get from a default.

## Several figures that have to match

An automatic choice depends on the data it sees, so two figures of different
data can end up on different scales. When they must match, there are two ways
round it.

**Fix the scale explicitly**, and pass the same values to every call:

```{r fixed}
map_surface(grid, "density", transform = "log", limits = c(0.001, 3),
            label = "animals per km2", coastline = land)
```

**Or let [`map_panels()`] do it**, which computes one scale over every panel
pooled and gives them a single collected legend:

```{r panels, fig.width = 8, fig.height = 3.2}
seasons <- cbind(spring = grid$density,
                 summer = grid$density * 2.5,
                 autumn = grid$density * 0.4)

map_panels(grid, seasons, label = "animals per km2", coastline = land,
           title = "Three periods on one scale")
```

This is worth insisting on. A per-panel scale makes each panel a picture of its
own relative pattern and then lays them out in a grid that invites reading
across -- so a quiet season and a busy one look identical, and the difference
between them, the only thing a series is for, is the one thing that has been
scaled away.

## The colours themselves

Three ramps, and they are checked rather than asserted. The package's test suite
simulates protanopia and deuteranopia and requires that the sequential and
bounded ramps keep **monotone luminance** under both -- so their ordering
survives when the hue difference does not -- and that the two arms of the
diverging ramp stay measurably separated.

```{r palettes, fig.height = 2}
fancymap_palette("sequential")

op <- par(mfrow = c(3, 1), mar = c(1, 0, 2, 0))
for (type in c("sequential", "bounded", "diverging")) {
  pal <- fancymap_palette(type, 32)
  image(matrix(seq_along(pal)), col = pal, axes = FALSE, main = type)
}
par(op)
```

The diverging ramp is blue against orange rather than the more usual blue
against red. Red–green confusion is the common deficiency, and a red arm
collapses toward the neutral middle under deuteranopia -- which is precisely the
failure that matters, since the whole job of a diverging ramp is to make "which
side of the centre" legible.

## The same scale, interactively

The `leaflet_*()` functions reuse these scale objects rather than re-deciding
anything, so a cell is the same colour in a static figure and an interactive
one. Handing a grid to `leaflet::colorNumeric()` directly would decide the
scale again at the call site -- linear, over the full data range, with no
capping -- and the two figures would show the same numbers in different colours.

```{r leaflet, eval = FALSE}
leaflet_surface(grid, "density", label = "animals per km2")
```

## In short

| the quantity | the function | what is decided for you | what you must decide |
|---|---|---|---|
| skewed, positive | `map_surface()` | transform, breaks, cap | nothing |
| bounded on \[0, 1\] | `map_probability()` | ends fixed at 0 and 1 | nothing |
| has a meaningful centre | `map_diverging()` | symmetric limits | **`midpoint`** |
| a series | `map_panels()` | one scale over all panels | nothing |

Everything in the third column is an argument, and every automatic choice is
reported when it is made. The point is not that these defaults are right for
every dataset. It is that they are written down, so a figure that needs
something else can say what and why.
